---
title: "JUMP Cell Painting cpg0016 CPCNN Embedding MAE v1.1"
author: "Michael Tran"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

This notebook demonstrates the structure and basic use of JUMP Cell Painting
cpg0016 CPCNN Embedding MAE v1.1, a `MultiAssayExperiment` containing well-level
Cell Painting embeddings and compound annotations.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(MultiAssayExperiment)
  library(SummarizedExperiment)
  library(dplyr)
  library(tibble)
  library(ggplot2)
  library(knitr)
  library(scales)
})

theme_set(theme_minimal(base_size = 12))

project_root <- normalizePath(
  if (dir.exists("data/procdata")) "." else "..",
  winslash = "/",
  mustWork = TRUE
)
knitr::opts_knit$set(root.dir = project_root)

mae_path <- file.path(project_root, "data", "procdata", "cpg0016_cpcnn_mae.rds")

if (!file.exists(mae_path)) {
  stop(
    "MAE file not found: ", mae_path
  )
}
```

## Load the MAE

```{r load-data}
mae <- readRDS(mae_path)
embedding_se <- experiments(mae)[["all_emb"]]
embedding_mat <- assay(embedding_se, "embedding")
dataset_name <- metadata(mae)$dataset_name
dataset_version <- metadata(mae)$dataset_version
dataset_label <- paste(dataset_name, dataset_version)

well_metadata <- as.data.frame(colData(mae)) |>
  rownames_to_column("sample_id") |>
  as_tibble()

compound_metadata <- as.data.frame(metadata(mae)$compound_metadata) |>
  as_tibble()

mae
```

## Dataset Overview

```{r dataset-overview}
perturbation_counts <- well_metadata |>
  count(Metadata_pert_type, name = "wells", sort = TRUE)

overview_tbl <- tibble(
  quantity = c(
    "Experiments",
    "Assay name",
    "Embedding dimensions",
    "Wells",
    "Plates",
    "Sources",
    "Unique compounds",
    "Treatment wells",
    "Negative-control wells",
    "Positive-control wells"
  ),
  value = c(
    length(experiments(mae)),
    paste(assayNames(embedding_se), collapse = ", "),
    scales::comma(nrow(embedding_mat)),
    scales::comma(ncol(embedding_mat)),
    scales::comma(n_distinct(well_metadata$Metadata_Plate)),
    scales::comma(n_distinct(well_metadata$Metadata_Source)),
    scales::comma(n_distinct(well_metadata$Metadata_JCP2022)),
    scales::comma(sum(well_metadata$Metadata_pert_type == "treatment")),
    scales::comma(sum(well_metadata$Metadata_pert_type == "negcon")),
    scales::comma(sum(well_metadata$Metadata_pert_type == "poscon"))
  )
)

kable(overview_tbl)
```

`r dataset_label` contains `r scales::comma(ncol(embedding_mat))` wells as
columns of a single embedding assay named `embedding`. Each well has a
`r scales::comma(nrow(embedding_mat))`-dimensional feature vector and a
matching row in `colData(mae)`.

## Inspect the MAE Structure

The object layout is inspected directly below. The experiment container, assay
name, dataset label, well-level metadata, and object-level metadata can all be accessed from
the printed structure and the code shown in this section.

```{r object-contents}
mae

experiments(mae)

metadata(mae)[c("dataset_name", "dataset_version", "compound_metadata_key")]

colnames(colData(mae))

names(metadata(mae))
```

## Well-Level Metadata

```{r well-metadata-head}
kable(head(well_metadata, 8))
```

The main fields commonly used in downstream analysis are:

* `sample_id`: the well identifier used throughout the object. It is the row name of `colData(mae)` and the column name of the embedding matrix.
* `Metadata_Source`: the source-specific study label.
* `Metadata_Batch`: the batch identifier within a source.
* `Metadata_Plate` and `Metadata_Well`: the physical plate and well coordinates.
* `Metadata_JCP2022`: the perturbagen identifier used to join to compound metadata, where the public alias `JUMP.CP.ID` is also available.
* `Metadata_InChIKey`: the structure-level identifier for the perturbagen.
* `Metadata_pert_type`: whether the well is a treatment, negative control (`negcon`), or positive control (`poscon`).

```{r perturbation-summary}
kable(perturbation_counts)
```

```{r perturbation-barplot, fig.height=4}
ggplot(
  perturbation_counts,
  aes(
    x = reorder(Metadata_pert_type, wells),
    y = wells,
    fill = Metadata_pert_type
  )
) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Well counts by perturbation type",
    x = "Perturbation type",
    y = "Number of wells"
  ) +
  coord_flip()
```

## Inspect the Embedding Assay

```{r rowdata-head}
kable(head(as.data.frame(rowData(embedding_se)), 3))
```

```{r rowdata-dimensions}
dim(rowData(embedding_se))
```

`all_emb` is a `SummarizedExperiment` with one assay called `embedding`. Rows are embedding features and columns are wells.

```{r embedding-dimensions}
tibble(
  metric = c("Embedding features per well", "Wells in assay"),
  value = c(nrow(embedding_mat), ncol(embedding_mat))
) |>
  kable()
```

```{r matrix-slice}
embedding_mat[1:6, 1:4]
```

The matrix and metadata are aligned by `sample_id`.

```{r alignment-check}
identical(colnames(embedding_mat), well_metadata$sample_id)
```

## Compound Metadata

```{r compound-metadata-head}
compound_preview <- compound_metadata |>
  select(
    `JUMP.CP.ID`,
    Metadata_JCP2022,
    InChIKey,
    Metadata_pert_type,
    Molecule.Name,
    Pubchem.CID,
    `In.JUMP.CP`,
    In_AnnotationDB
  ) |>
  slice_head(n = 8)

kable(compound_preview)
```

This table is stored once at the object level instead of being repeated on
every well. The public join key is `JUMP.CP.ID`, and the compatibility alias
`Metadata_JCP2022` is retained to simplify joins to the well-level metadata.

The `Metadata_pert_type` labels also appear in the compound metadata. In this
context, `negcon` denotes negative controls and `poscon` denotes positive
controls.

```{r control-compound-examples}
negcon_examples <- compound_metadata |>
  filter(Metadata_pert_type == "negcon") |>
  select(
    `JUMP.CP.ID`,
    Metadata_Control_Name,
    Molecule.Name,
    Pubchem.CID
  ) |>
  slice_head(n = 3)

poscon_examples <- compound_metadata |>
  filter(Metadata_pert_type == "poscon") |>
  select(
    `JUMP.CP.ID`,
    Metadata_Control_Name,
    Molecule.Name,
    Pubchem.CID
  ) |>
  slice_head(n = 3)

kable(negcon_examples, caption = "Example negative-control compounds")

kable(poscon_examples, caption = "Example positive-control compounds")
```

## Demonstrate Basic Access Patterns

### Pull One Plate and Its Embeddings

```{r plate-subset}
plate_id <- well_metadata$Metadata_Plate[1]

plate_wells <- well_metadata |>
  filter(Metadata_Plate == plate_id)

plate_embeddings <- embedding_mat[, plate_wells$sample_id, drop = FALSE]

kable(
  plate_wells |>
    select(
      sample_id,
      Metadata_Well,
      Metadata_JCP2022,
      Metadata_pert_type
    ) |>
    slice_head(n = 8)
)

dim(plate_embeddings)
```

### Annotate Wells with Compound Metadata

```{r plate-annotations}
plate_treatment_wells <- plate_wells |>
  filter(Metadata_pert_type == "treatment")

plate_treatment_annotated <- plate_treatment_wells |>
  transmute(
    sample_id,
    Metadata_Source,
    Metadata_Plate,
    Metadata_Well,
    `JUMP.CP.ID` = Metadata_JCP2022
  ) |>
  left_join(
    compound_metadata |>
      select(
        `JUMP.CP.ID`,
        Molecule.Name,
        Pubchem.CID,
        In_AnnotationDB
      ),
    by = "JUMP.CP.ID"
  )

kable(slice_head(plate_treatment_annotated, n = 8))
```

The same join pattern works for any well subset:

```{r full-treatment-pattern, eval=FALSE}
treatment_wells <- well_metadata |>
  filter(Metadata_pert_type == "treatment")

treatment_embeddings <- embedding_mat[, treatment_wells$sample_id, drop = FALSE]

treatment_annotations <- treatment_wells |>
  transmute(
    sample_id,
    Metadata_Source,
    Metadata_Plate,
    Metadata_Well,
    `JUMP.CP.ID` = Metadata_JCP2022
  ) |>
  left_join(
    compound_metadata,
    by = "JUMP.CP.ID"
  )
```

### Extract Embeddings for a Small Well Set

```{r small-well-set}
example_wells <- plate_treatment_wells$sample_id[1:4]

embedding_mat[1:10, example_wells, drop = FALSE]
```

This gives a small feature-by-well slice. For modeling or distance
calculations, the wells are typically subset first and the result is then kept
as a matrix.

### Build One Mean Profile Per Compound

```{r compound-mean-pattern, eval=FALSE}
treatment_wells <- well_metadata |>
  filter(Metadata_pert_type == "treatment")

treatment_embeddings <- embedding_mat[, treatment_wells$sample_id, drop = FALSE]

compound_means <- rowsum(
  x = t(treatment_embeddings),
  group = treatment_wells$Metadata_JCP2022,
  reorder = FALSE
)

compound_counts <- table(treatment_wells$Metadata_JCP2022)

compound_means <- compound_means /
  as.numeric(compound_counts[rownames(compound_means)])

dim(compound_means)
```

After this step, `compound_means` has one row per compound and one column per
embedding feature.

## Session Info

```{r session-info}
sessionInfo()
```
