---
title: "Cell Painting cpg0016-jump Embedding MAE QC"
author: "Michael Tran"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

This notebook demonstrates the structure and basic use of the final
`MultiAssayExperiment` containing well-level Cell Painting embeddings,
CellProfiler profiles, and compound annotations.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(MultiAssayExperiment)
  library(SummarizedExperiment)
  library(dplyr)
  library(tibble)
  library(ggplot2)
  library(knitr)
  library(scales)
  library(uwot)
})

theme_set(theme_minimal(base_size = 12))

project_root <- normalizePath(
  if (dir.exists("data/results")) "." else "..",
  winslash = "/",
  mustWork = TRUE
)
knitr::opts_knit$set(root.dir = project_root)

mae_path <- file.path(
  project_root,
  "data",
  "results",
  "cpg0016_emb_MultiAssayExperiment.rds"
)

if (!file.exists(mae_path)) {
  stop("MAE file not found: ", mae_path)
}
```

## Load the MAE

```{r load-data}
mae <- readRDS(mae_path)

assay_names <- names(experiments(mae))
example_assay <- "cpcnn"
embedding_se <- experiments(mae)[[example_assay]]
embedding_mat <- assay(embedding_se)
cellprofiler_se <- experiments(mae)[["cellprofiler"]]
cellprofiler_mat <- assay(cellprofiler_se)

well_metadata <- as.data.frame(colData(mae)) |>
  as_tibble()

compound_metadata <- as.data.frame(metadata(mae)$Drug.Metadata) |>
  as_tibble()

mae
```

## Dataset Overview

```{r dataset-overview, class.source="fold-hide"}
assay_summary <- tibble(
  assay = assay_names,
  feature_type = case_when(
    assay_names == "cpcnn" ~ "embedding vector positions",
    assay_names == "cellprofiler" ~ "scalar morphology features",
    TRUE ~ "numeric profile features"
  ),
  assay_name = vapply(
    assay_names,
    \(assay_name) {
      paste(assayNames(experiments(mae)[[assay_name]]), collapse = ", ")
    },
    character(1)
  ),
  features = vapply(
    assay_names,
    \(assay_name) nrow(experiments(mae)[[assay_name]]),
    integer(1)
  ),
  wells = vapply(
    assay_names,
    \(assay_name) ncol(experiments(mae)[[assay_name]]),
    integer(1)
  ),
  profile_model = vapply(
    assay_names,
    \(assay_name) {
      paste(
        unique(rowData(experiments(mae)[[assay_name]])$Profile.Model),
        collapse = ", "
      )
    },
    character(1)
  )
)

perturbation_counts <- well_metadata |>
  count(Perturbation.Type, name = "wells", sort = TRUE)

assay_overview_tbl <- bind_rows(
  assay_summary |>
    transmute(
      quantity = paste(assay, "features"),
      value = scales::comma(features)
    ),
  assay_summary |>
    transmute(
      quantity = paste(assay, "wells"),
      value = scales::comma(wells)
    )
)

overview_tbl <- tibble(
  quantity = c(
    "Pipeline ID",
    "Pipeline version",
    "Experiments",
    "Assays",
    "Wells in colData",
    "Plates",
    "Sources",
    "Unique compounds",
    "Drug metadata records",
    "Treatment wells",
    "Negative-control wells",
    "Positive-control wells"
  ),
  value = c(
    metadata(mae)$Pipeline$ID,
    metadata(mae)$Pipeline$Version,
    scales::comma(length(experiments(mae))),
    paste(assay_names, collapse = ", "),
    scales::comma(nrow(well_metadata)),
    scales::comma(n_distinct(well_metadata$Plate)),
    scales::comma(n_distinct(well_metadata$Source)),
    scales::comma(n_distinct(well_metadata$JUMP.CP.ID)),
    scales::comma(nrow(compound_metadata)),
    scales::comma(sum(well_metadata$Perturbation.Type == "treatment")),
    scales::comma(sum(well_metadata$Perturbation.Type == "negcon")),
    scales::comma(sum(well_metadata$Perturbation.Type == "poscon"))
  )
) |>
  bind_rows(assay_overview_tbl)

kable(overview_tbl)
```

```{r assay-summary}
kable(assay_summary)
```

This file contains `r length(experiments(mae))` assays. The `cpcnn` assay stores
`r scales::comma(nrow(embedding_mat))` embedding features across
`r scales::comma(ncol(embedding_mat))` wells. The `cellprofiler` assay stores
scalar morphology features for the wells where CellProfiler profiles were
available.

## Inspect the MAE Structure

The object layout is inspected directly below. The experiment container, assay
names, well-level metadata, sample map, and object-level metadata can all be
accessed from the printed structure and the code shown in this section.

```{r object-contents}
mae

experiments(mae)

colnames(colData(mae))

head(sampleMap(mae))

names(metadata(mae))
```

## Well-Level Metadata

```{r well-metadata-head}
well_metadata |>
  select(
    Sample.ID,
    JUMP.CP.ID,
    Pubchem.CID,
    InChIKey,
    Source,
    Batch,
    Plate,
    Well,
    Perturbation.Type
  ) |>
  slice_head(n = 8) |>
  kable()
```

The main fields commonly used in downstream analysis are:

* `Sample.ID`: the well identifier used throughout the object. It is the row name of `colData(mae)` and the column name used by assay matrices.
* `Source`: the source-specific study label.
* `Batch`: the batch identifier within a source.
* `Plate` and `Well`: the physical plate and well coordinates.
* `JUMP.CP.ID`: the perturbagen identifier used to join to compound metadata.
* `InChIKey`: the structure-level identifier for the perturbagen.
* `Perturbation.Type`: whether the well is a treatment, negative control (`negcon`), or positive control (`poscon`).

```{r perturbation-summary}
kable(perturbation_counts)
```

```{r perturbation-barplot, fig.height=4}
ggplot(
  perturbation_counts,
  aes(
    x = reorder(Perturbation.Type, wells),
    y = wells,
    fill = Perturbation.Type
  )
) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Well counts by perturbation type",
    x = "Perturbation type",
    y = "Number of wells"
  ) +
  coord_flip()
```

Treatment compound replication is summarized below. This view focuses on
treatments; controls are represented separately in the perturbation summaries.

```{r treatment-compound-replicates, fig.height=5}
compound_replicates <- well_metadata |>
  filter(Perturbation.Type == "treatment") |>
  count(JUMP.CP.ID, name = "wells", sort = TRUE) |>
  left_join(
    compound_metadata |>
      select(
        JUMP.CP.ID,
        AnnotationDB.Name,
        Pubchem.CID,
        InChIKey,
        Control.Name
      ),
    by = "JUMP.CP.ID"
  )

replicate_summary_tbl <- tibble(
  quantity = c(
    "Treatment compounds",
    "Treatment compounds with more than one well",
    "Median wells per treatment compound",
    "95th percentile wells per treatment compound",
    "Maximum wells for one treatment compound"
  ),
  value = c(
    scales::comma(nrow(compound_replicates)),
    scales::comma(sum(compound_replicates$wells > 1)),
    scales::comma(median(compound_replicates$wells)),
    scales::comma(quantile(compound_replicates$wells, 0.95, names = FALSE)),
    scales::comma(max(compound_replicates$wells))
  )
)

kable(replicate_summary_tbl)

ggplot(
  compound_replicates,
  aes(x = wells)
) +
  geom_histogram(bins = 45, fill = "#3B6EA8", color = "white") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Treatment wells per compound",
    subtitle = "Controls are excluded by perturbation type.",
    x = "Wells per treatment compound",
    y = "Compounds"
  )
```

## Inspect the CPCNN Assay

```{r rowdata-head}
kable(head(as.data.frame(rowData(embedding_se)), 3))
```

```{r rowdata-dimensions}
dim(rowData(embedding_se))
```

The `cpcnn` experiment is a `SummarizedExperiment` with one assay called
`values`. Rows are embedding features and columns are wells.

```{r embedding-dimensions}
tibble(
  metric = c("Embedding features per well", "Wells in assay"),
  value = c(nrow(embedding_mat), ncol(embedding_mat))
) |>
  kable()
```

The UMAP below is a visualization of a reproducible sample of 1,000 treatment
wells from the `cpcnn` assay. Each point is one well-level profile.

```{r cpcnn-well-umap, fig.height=5.5, fig.width=7}
set.seed(20260429)

cpcnn_treatment_wells <- well_metadata |>
  filter(
    Perturbation.Type == "treatment",
    Sample.ID %in% colnames(embedding_mat)
  ) |>
  left_join(
    compound_metadata |>
      select(JUMP.CP.ID, Control.Name),
    by = "JUMP.CP.ID"
  ) |>
  filter(is.na(Control.Name))

umap_sample_n <- min(1000L, nrow(cpcnn_treatment_wells))

umap_wells <- cpcnn_treatment_wells |>
  slice_sample(n = umap_sample_n)

umap_embeddings <- embedding_mat[, umap_wells$Sample.ID, drop = FALSE]

umap_input <- scale(t(umap_embeddings))
umap_input[is.na(umap_input)] <- 0

umap_layout <- uwot::umap(
  umap_input,
  n_neighbors = 15,
  min_dist = 0.2,
  metric = "cosine",
  n_threads = 1,
  n_sgd_threads = 1,
  verbose = FALSE
)

umap_tbl <- as_tibble(umap_layout, .name_repair = \(x) c("UMAP1", "UMAP2")) |>
  bind_cols(
    umap_wells |>
      select(Sample.ID, JUMP.CP.ID, Source, Plate, Well)
  )

ggplot(
  umap_tbl,
  aes(x = UMAP1, y = UMAP2, color = Source)
) +
  geom_point(alpha = 0.65, size = 1.3) +
  labs(
    title = "UMAP of sampled cpcnn well profiles",
    subtitle = paste0(
      "Well-level embeddings for ",
      scales::comma(nrow(umap_tbl)),
      " sampled treatment wells; controls excluded."
    ),
    x = "UMAP 1",
    y = "UMAP 2",
    color = "Source"
  )
```

```{r matrix-slice}
embedding_mat[1:6, 1:4]
```

The `cpcnn` assay columns are represented in `colData(mae)`.

```{r alignment-check}
all(colnames(embedding_mat) %in% well_metadata$Sample.ID)
```

## Inspect the CellProfiler Assay

The `cellprofiler` experiment has the same `SummarizedExperiment` layout as
`cpcnn`: one `values` assay with features as rows and wells as columns. The
main difference is feature meaning. `cpcnn` rows are embedding vector
positions, while `cellprofiler` rows are scalar morphology features from the
CellProfiler profile table.

```{r cellprofiler-rowdata-head}
kable(head(as.data.frame(rowData(cellprofiler_se)), 8))
```

```{r cellprofiler-summary}
tibble(
  metric = c(
    "CellProfiler features per well",
    "CellProfiler wells in assay",
    "CellProfiler wells represented in colData",
    "CellProfiler wells not in colData"
  ),
  value = c(
    nrow(cellprofiler_mat),
    ncol(cellprofiler_mat),
    sum(colnames(cellprofiler_mat) %in% well_metadata$Sample.ID),
    sum(!(colnames(cellprofiler_mat) %in% well_metadata$Sample.ID))
  )
) |>
  kable()
```

```{r cellprofiler-matrix-slice}
cellprofiler_mat[1:6, 1:4]
```

## Compound Metadata

```{r compound-metadata-head}
compound_preview <- compound_metadata |>
  select(
    JUMP.CP.ID,
    InChIKey,
    Perturbation.Type,
    AnnotationDB.Name,
    Pubchem.CID,
    In.AnnotationDB
  ) |>
  slice_head(n = 8)

kable(compound_preview)
```

This table is stored once at the object level instead of being repeated on
every well. The join key is `JUMP.CP.ID`.

The `Perturbation.Type` labels also appear in the compound metadata. In this
context, `negcon` denotes negative controls and `poscon` denotes positive
controls.

For compounds represented in the `cpcnn` assay, most records are enriched with
AnnotationDB names and identifiers. Some compounds, such as `JCP2022_000004`,
do not have mapped AnnotationDB names, PubChem CIDs, or AnnotationDB SMILES.
The overall breakdown is shown below.

```{r annotationdb-availability, fig.height=4.5}
cpcnn_compound_metadata <- well_metadata |>
  filter(Sample.ID %in% colnames(embedding_mat)) |>
  distinct(JUMP.CP.ID) |>
  left_join(compound_metadata, by = "JUMP.CP.ID")

annotation_availability <- cpcnn_compound_metadata |>
  mutate(
    annotation_status = if_else(
      In.AnnotationDB,
      "AnnotationDB enriched",
      "No AnnotationDB match",
      missing = "No AnnotationDB match"
    )
  ) |>
  count(annotation_status, name = "compounds") |>
  mutate(
    fraction = compounds / sum(compounds),
    percent = scales::percent(fraction, accuracy = 0.1),
    label = paste0(
      annotation_status,
      "\n",
      scales::comma(compounds),
      " (",
      percent,
      ")"
    )
  )

kable(
  annotation_availability |>
    select(annotation_status, compounds, percent)
)

ggplot(
  annotation_availability,
  aes(x = "", y = fraction, fill = annotation_status)
) +
  geom_col(width = 1, color = "white") +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    lineheight = 0.95
  ) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#4C956C", "#C75C5C")) +
  labs(
    title = "AnnotationDB metadata availability for cpcnn compounds",
    fill = NULL
  ) +
  theme_void()
```

```{r control-compound-examples}
negcon_examples <- compound_metadata |>
  filter(Perturbation.Type == "negcon") |>
  select(
    JUMP.CP.ID,
    Control.Name,
    AnnotationDB.Name,
    Pubchem.CID
  ) |>
  slice_head(n = 3)

poscon_examples <- compound_metadata |>
  filter(Perturbation.Type == "poscon") |>
  arrange(desc(JUMP.CP.ID)) |>
  select(
    JUMP.CP.ID,
    Control.Name,
    AnnotationDB.Name,
    Pubchem.CID
  ) |>
  slice_head(n = 3)

kable(negcon_examples, caption = "Example negative-control compounds")

kable(poscon_examples, caption = "Example positive-control compounds")
```

## Demonstrate Basic Access Patterns

### Pull One Plate and Its Embeddings

```{r plate-subset}
plate_id <- well_metadata$Plate[1]

plate_wells <- well_metadata |>
  filter(Plate == plate_id, Sample.ID %in% colnames(embedding_mat))

plate_embeddings <- embedding_mat[, plate_wells$Sample.ID, drop = FALSE]

kable(
  plate_wells |>
    select(
      Sample.ID,
      Well,
      JUMP.CP.ID,
      Perturbation.Type
    ) |>
    slice_head(n = 8)
)

dim(plate_embeddings)
```

### Annotate Wells with Compound Metadata

```{r plate-annotations}
plate_treatment_wells <- plate_wells |>
  filter(Perturbation.Type == "treatment")

plate_treatment_annotated <- plate_treatment_wells |>
  select(
    Sample.ID,
    Source,
    Plate,
    Well,
    JUMP.CP.ID
  ) |>
  left_join(
    compound_metadata |>
      select(
        JUMP.CP.ID,
        AnnotationDB.Name,
        Pubchem.CID,
        In.AnnotationDB
      ),
    by = "JUMP.CP.ID"
  )

kable(slice_head(plate_treatment_annotated, n = 8))
```

The same join pattern works for any well subset:

```{r full-treatment-pattern, eval=FALSE}
treatment_wells <- well_metadata |>
  filter(
    Perturbation.Type == "treatment",
    Sample.ID %in% colnames(embedding_mat)
  )

treatment_embeddings <- embedding_mat[, treatment_wells$Sample.ID, drop = FALSE]

treatment_annotations <- treatment_wells |>
  left_join(
    compound_metadata,
    by = "JUMP.CP.ID"
  )
```

### Extract Embeddings for a Small Well Set

```{r small-well-set}
example_wells <- plate_treatment_wells$Sample.ID[seq_len(min(
  4,
  nrow(plate_treatment_wells)
))]

embedding_mat[1:10, example_wells, drop = FALSE]
```

This gives a small feature-by-well slice. For modeling or distance
calculations, the wells are typically subset first and the result is then kept
as a matrix.

### Build One Mean Profile Per Compound

```{r compound-mean-pattern}
treatment_wells <- well_metadata |>
  filter(
    Perturbation.Type == "treatment",
    Sample.ID %in% colnames(embedding_mat)
  )

example_compounds <- treatment_wells |>
  count(JUMP.CP.ID, name = "n_wells", sort = TRUE) |>
  filter(n_wells > 1) |>
  slice_head(n = 5)

example_treatment_wells <- treatment_wells |>
  semi_join(example_compounds, by = "JUMP.CP.ID")

treatment_embeddings <- embedding_mat[,
  example_treatment_wells$Sample.ID,
  drop = FALSE
]

compound_means <- rowsum(
  x = t(treatment_embeddings),
  group = example_treatment_wells$JUMP.CP.ID,
  reorder = FALSE
)

compound_counts <- table(example_treatment_wells$JUMP.CP.ID)

compound_means <- compound_means /
  as.numeric(compound_counts[rownames(compound_means)])

compound_mean_preview <- as.data.frame(
  compound_means[, seq_len(min(6, ncol(compound_means))), drop = FALSE],
  check.names = FALSE
) |>
  rownames_to_column("JUMP.CP.ID") |>
  as_tibble() |>
  left_join(example_compounds, by = "JUMP.CP.ID") |>
  select(JUMP.CP.ID, n_wells, everything())

kable(compound_mean_preview)

dim(compound_means)
```

This preview averages the `cpcnn` embeddings for a few replicated treatment
compounds. After this step, `compound_means` has one row per compound and one
column per embedding feature.

## Session Info

```{r session-info}
sessionInfo()
```
